---
title: "emgR tutorials"
author: "Yusuke Matsui"
date: "1/2/2020"
output: ioslides_presentation
---

## `emgR` overview
- Analysis suites for surface electromyography (sEMG) data
- Basic process functionality
    - Bandpass filter
    - Qunatification
    - Normalization
- Advanced analysis
    - Waveform simililarity analysis
        - Coherence analysis (incl. wavelet)
        - Auto/cross correlation analysis
    - Similiarity based muscle network analysis

## Install emgR package
```{r,warning=FALSE,message=F}
# install.packages("devtools") if not installed
 devtools::install_github("https://github.com/ymatts/emgR")
#install.packages("~/Dropbox/matsui/git/emgR.tar.gz",repos = NULL,type = "source")
library(emgR)
```

## Data acquisition
- From [Santuz et al. (2018)](https://www.frontiersin.org/articles/10.3389/fphys.2018.01509/full)
- 13 muscles sEMG (1000 Hz) records during gait cycle
- Time data will be required for constructing emgR object
```{r,warning=F,message=F}
data(emg)
head(emg,3)
```

## Constructing object
- `emgList` is a method for constructing emgR object
- `emgList` object is simply a `list` data structure
-  Input time (sec) and multi channel sEMG data
    - Time is in row and Muscles are in colums

```{r,warning=TRUE,message=FALSE}
tt = emg$Time # Time record
xx = emg[,-1] # EMG record
obj = emgR::emgList(x = xx,t = tt)
obj
```

## Add annotation
- Annotation file must be following format
    - `id` in first column may indicate one cycle or trial
    - `type` in second column may indicate description of id
    - `start` and `end` should be annotated/triggered time in second
```{r}
(f = system.file("extdata","gaitAnno.txt",package = "emgR"))
head(read.table(f,header = T,sep="\t"),3)
```

## Add annotation
- `addAnnotation` function adds indices in raw emg records with annotated time
```{r}
obj = emgR::addAnnotation(obj,annoFile = f)
head(obj$annotation,3)
```

## Extract EMG signals followed by annotation
```{r}

```

## Filtering sEMG signals
- Bandpass filter using butterworth filter
- Filtered objects will be overwride in `emgR`
- Default setting is `order` 4 and `[5,200]` bandpass filter
```{r,echo=T,warning=TRUE,message=FALSE}
obj = emgR::btw(obj,order = 4,filtFreq = c(5,200))
obj
```

## Normalize sEMG signals
- Several normalizing methods will be implemted
    - Default setting use simple Maximum Voluntary Contruct (MVC)
```{r,echo=T,warning=F,message=F}
obj = emgR::normalize(obj)
obj
```

## Extract gait cycle
- 41 gait cycles
- Each line corresponds to a cycle
    - First column is times during heel contact
    - Second column is times during stance
```{r,warning=F,message=F}
load("data/cycleTime.rda")
head(cycleTime)
```



## Waveform similiarity analysis
- Coherence analysis
    - Linear coherence / Wavelet coherence
- Correlation analysis
    - Auto/Cross-correlation, Distance correlation etc.
- Default setting is linear coherence by `stats::spec.pgram()` function
```{r,echo=T,warning=F,message=F}
obj = emgR::coherence(obj)
obj
```

## Visualize coherence function
- Muscle $^times$ Muscle smoothed coherences are plotted
- The plot is saved in file path with `files`
- `cohRange` is which frequency range should be considered
- This takes several time
```{r,eval=F,warning=F,message=F}
file_path = tempfile()
plot(obj,type="coh",files = file_path,cohRange=c(0,100))
```

## Visualize coherence functions
```{r,echo=F,warning=F,message=F,out.width="70%",fig.align="center"}
knitr::include_graphics(file_path)
```

## Aggregate coherence function into means within frequency bands
- Average aggregation by frequecies such as $\alpha$, $\beta$, $\gamma$ bands.
- Frequency can be specified with `cohRange` by vector.
    - Following example shows averaged coherence from 15 Hz to 20 Hz. 
- Heatmap of aggregated coherence can be visulized by `plot.it` argument
```{r,warning=F,message=F}
obj = emgR::aggrCoh(obj,cohRange = c(15,20),plot.it = FALSE)
obj
```

## Aggregate coherence function into means within frequency bands
```{r,warning=F,message=F,out.width="90%",fig.align="center"}
obj = emgR::aggrCoh(obj,cohRange = c(15,20),plot.it = TRUE)
```


## Coherent muscle network construction
- Network is constructed via aggregated coherence
- Two mode of networks:
    - Weighted network can be constructed with `mode="weighted"`
    - Not weighted network can be constructed with `mode="bin"`
- Coherence is binarized with `thres` argument
    - Default `thres` is `0.25`
- You can visualize network via `plot.it` argument.
```{r,warning=F,message=F}
obj = emgR::cohNet(obj,mode = "bin",thres = 0.25,plot.it = FALSE)
```

## Coherent muscle network construction
```{r,warning=F,message=F,out.width="90%",fig.align="center"}
obj = emgR::cohNet(obj,mode = "bin",thres = 0.25,plot.it = TRUE)
```

## Network based muscle module detection
- Muscle modules are found based on linkcomm pakcage ([Ahn et al. 2010]("https://arxiv.org/pdf/0903.3178.pdf"), [Becker et al. 2012]("http://bioinformatics.oxfordjournals.org/content/28/1/84.full"))
- `plot.it` visualize module structure
    - You can choose `plot.type` from `"member"` or `"graph"`
- In this model, each muscle can belong to several modules (muti-membership)


## Network based muscle module detection
- In case of `plot.type = "graph"`
```{r,warning=F,message=F}
obj = emgR::moduleFinder(obj,plot.it = TRUE,plot.type = "graph")
```

## Network based muscle module detection
- In case of `plot.type = "member"`
```{r,warning=F,message=F}
obj = emgR::moduleFinder(obj,plot.it = TRUE,plot.type = "member")
```


## Muscle synergy (2D-array)
- Derive muscle synergy based on Non-negative matrix factorzation (NMF) algorithm


